---
title: "Regression Models - MPG of Manual vs. Auto in the MTCARS dataset"
author: "Michael Bien"
date: "`r Sys.Date()`"
output: 
  html_document:
    fig_caption: yes 
    keep_md: yes
---

# Executive Summary
In the analysis below, I examine the MTCARS dataset to determine the relationship between a manual transmission and MPG.  Specifically, I answer these two questions: 

1. "Is an automatic or manual better for MPG?"
2. "Quantify the MPG difference between automatic and manual transmissions."

**Conclusion**

While the initial observational summary shows that cars with a manual transmission get better MPG (+7.2 MPG for manual), a multivariate model reveals a more complex relationship, when considering other explanatory variables like a vehicle's weight.  **Ultimately, I conclude that an manual transmission gets slightly worse MPG (-.02MPG for manual) when controlling for the vehicle's weight.**

**Methodology**
As shown below, first, I consider whether the fits a normal distribution.  Second, perform EDA to explore the relationships between key variables, and identify potential predictor variables.  Then, I build several models considering the effect of a manual transmission on MPG.  I build the following 4 models, and then choose the second model based on results from the ANOVA function, Variance Inflation results, and model performance (R squared) metrics.

1. Model1: mpg ~ transmission  
**2. Model2: mpg ~ wt + transmission  **  **<- Selected for low p value, low VIF, and high R^2 **
3. Model3: mpg ~ wt * transmission  
4. Model4: mpg ~ wt + transmission + disp + hp  

## Setup
Start by setting (echo = TRUE) and declaring libraries to be used
```{r setup, include=TRUE, message=FALSE, cache = TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)  #"always use echo = TRUE" for this assignment, per notes
library(ggplot2)
library(regclass)
library(Hmisc)
```

## Data Import and Cleaning

In this section, I import the data and ensure that it follows a normal distribution so that lm() can be used.

```{r data_step, cache = TRUE, message=FALSE}
data(mtcars)
#str(mtcars)
## mpg, transmission, and several other potential variables

  ###########################################################
  ###                  Recode as factors                  ###
  ###########################################################
  mtcars$transmission <- as.factor(ifelse(mtcars$am == 0, "automatic", "manual"))
  mtcars$cyl <- as.factor(mtcars$cyl)
  mtcars$vs <- as.factor(mtcars$vs)


#crosstable to confirm
  with(mtcars, table(am, transmission))
  
  ###########################################################
  ###             Are the residuals normal?               ###
  ###########################################################
  
  qqnorm(mtcars$mpg, pch=1, main="QQPLOT of mtcars$mpg")
  #linear model is appropriate for this case

```

## Exploratory data analysis

In this section, I create a boxplot of MPG by manual vs. automatic, noting that manual gets better mileage; however, faceting by weight, this effect diminishes.

```{r EDA, cache = TRUE, message=FALSE}

  ###########################################################
  ###             Exploratory Data Analysis               ###
  ###########################################################
  
  
  # 1 way look - MPG of manual vs. auto
  # Average MPG: 17 for automatic, 24 for manual
  round(tapply(mtcars$mpg, mtcars$transmission, mean),1)
  
  #In this plot, we note that manual cars often get better MPG

  ggplot(data=mtcars, aes(x=transmission, y=mpg)) + 
    geom_boxplot()+
    geom_jitter() + 
    ggtitle("MPG of auto vs. manual")
  
  # #facet plots to examine average MPG within other variables
  # 
  # ggplot(data=mtcars, aes(x=transmission, y=mpg)) + 
  #   geom_boxplot()+
  #   facet_wrap(~cyl)

#cut the weight variable
  summary(mtcars$wt)
  #hist(mtcars$wt)
  #Cut the weight variable into 5 groups
  mtcars$weight_group <- cut(mtcars$wt, 5)
  #summary(mtcars$weight_group)
  
  #In this plot, we note that, when controlling for weight, automatic and manual perform similarly
  
  ggplot(data=mtcars, aes(x=transmission, y=mpg)) + 
    geom_boxplot()+
    facet_wrap(~weight_group) + 
    geom_jitter() + 
    ggtitle("MPG of auto vs. manual by weight")
  
  vars_to_drop <- c("am", "drat","displ", "qsec", "vs", "gear", "carb", "weight_group", "cyl")
  mtcars2 <- mtcars[,!names(mtcars) %in% vars_to_drop]
  
  pairs(mpg  ~ . , mtcars2)

```
  
  
## Model Fitting 

In this section, I fit several models and, using ANOVA, VIF, and R squared as metrics, select a model

```{r model_fitting_step, cache = TRUE, message=FALSE}

  # manual by itself
  model1 <- lm(mpg ~ transmission, mtcars2)
  summary(model1)$coefficients
  
  #slightly negative when we take into account weight
  model2 <- lm(mpg ~ wt + transmission, mtcars2)
  model3 <- lm(mpg ~ wt * transmission, mtcars2)
  model4 <- lm(mpg ~ wt + transmission + disp + hp, mtcars2)

  ###########################################################
  ###             Model Selection, including VIF          ###
  ###########################################################  
  
  anova(model1, model2, model3, model4)
  
  #Based on the ANOVA above, models 3 and 2 are potential predictors based on their p values; model 4 is excluded

  summary(model1)$r.squared
  summary(model2)$r.squared
  summary(model3)$r.squared
  
  VIF(model2)
  VIF(model3)
  
  #Model 3 has 2 high a VIF to be considered; model 2 is selected
  
  summary(model2)$coefficients
  initial_coef <- summary(model1)$coefficients
  final_coefs <- summary(model2)$coefficients
  
  #Since anova excludes model 4 and VIF excludes model 3, I note that model 2 yields a parsimonious and trustworthy model.


```

However, it should be pointed out that there is some uncertainty here, as the R^2 of .75 only explains 75% of the variance.


## Residuals

Finally, I look at the model fit residuals to ensure our fit was appropriate and see no major issues.

```{r residuals, cache = TRUE, message=FALSE}
par(mfrow=c(2,2))    # set the plotting area into a 2*2 array
plot(model2)


```

Final model coefficients: `r final_coefs`